### 2025-05-23 docs/article/Sections/problem-statement.tex
Chapter 3: start with reactive approaches in internal DSLs, contrast with synchronous external DSLs, introduce virtual threads as possible solution.

In this section we discuss the problems associated with progressive server-side
rendering (PSSR) in modern web applications, focusing on the challenges and
limitations of existing approaches. We then present the objective of our work,
which aims to increase the available options for PSSR in web applications,
particularly in the context of JVM based frameworks.

\textit As we outlined in the previous section, \textit{external} DSLs for HTML such as Thymeleaf, Jstachio, Handlebars, and others define HTML
templates within HTML documents with markers specific to the engine. Template engines that use \textit{internal} DSLs, such as KotlinX or HtmlFlow,
define the HTML templates in the programming language itself, leveraging the language's
syntax and features to create the templates.

\textit{Internal} DSLs can more easily support asynchronous data models since
they can leverage the language's existing features and libraries, thus enabling
the progressive streaming of HTML content to the client, with \textit{external}
DSLs supporting simpler blocking data models such as \texttt{Iterable} or
\texttt{List}. While on the previous section we outlined several advantages of
using \textit{internal} DSLs over textual templates for HTML, there are also
advantages to using \textit{external} DSLs:

\begin{enumerate}
    \item \emph{Separation of Concerns}: \textit{External} DSLs allow for a clear separation between
          the HTML structure and the application logic, making it easier to maintain and update
          templates without modifying the underlying code. This ''
          separation'' can be beneficial, as it allows front-end developers to work on the HTML
          templates without needing to understand the underlying application logic or programming language.
    \item \emph{Cross-Language Compatibility}: \textit{External} DSLs can be used across different programming languages
          and frameworks, making them more versatile and easier to integrate into existing projects.
          This is particularly useful in environments where multiple languages are used.
    \item \emph{Familiarity}: Many developers are already familiar with HTML and its syntax, making
          it easier to work with \textit{external} DSLs that use HTML-like syntax.
\end{enumerate}

Due to these advantages, many developers prefer to use \textit{external} DSLs
for HTML templates, even if it means sacrificing some of the advantages of
using \textit{internal} DSLs. However, this preference can lead to challenges
when it comes to progressive server-side rendering (PSSR) and non-blocking I/O,
as many \textit{external} DSLs do not support asynchronous data models.

\autoref{lst:presentation-jstachio} shows the HTML template for a \texttt{Presentation} class using the JStachio template engine, 
while \autoref{lst:presentation-htmlflow} presents the equivalent template using HtmlFlow.

\lstset{style=listingstyle}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{JStachio}}
  label={lst:presentation-jstachio}
]
<!DOCTYPE html>
<html lang="en-US">
<head>
    <meta charset="utf-8">
    <meta name="viewport" content="width=device-width, initial-scale=1.0">
    <meta http-equiv="X-UA-compatible" content="IE=Edge">
    <title>
        JFall 2013 Presentations - JStachio
    </title>
    <link rel="Stylesheet" href="/webjars/bootstrap/5.3.0/css/bootstrap.min.css">
</head>
<body>
<div class="container">
    <div class="pb-2 mt-4 mb-3 border-bottom">
        <h1>
            JFall 2013 Presentations - JStachio
        </h1>
    </div>
    {{#presentationItems}}
    <div class="card mb-3 shadow-sm rounded">
        <div class="card-header">
            <h5 class="card-title">
                {{title}} - {{speakerName}}
            </h5>
        </div>
        <div class="card-body">
            {{summary}}
        </div>
    </div>
    {{/presentationItems}}
</div>
</body>
</html>
\end{lstlisting}
\vspace{0.5cm}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{HtmlFlow}}
  label={lst:presentation-htmlflow}
]
 val htmlFlowTemplate: HtmlViewAsync<Observable<Presentation>> =
    HtmlFlow.viewAsync<Observable<Presentation>?> { view ->
        view.html {
            attrLang("en-us")
            head {
                meta { attrCharset("UTF-8") }
                meta { attrName("viewport").attrContent("width=device-width, initial-scale=1.0") }
                meta { addAttr("http-equiv", "X-UA-Compatible").attrContent("IE=Edge") }
                title { text("JFall 2013 Presentations - HtmlFlow") }
                link { attrRel(EnumRelType.STYLESHEET).attrHref("/webjars/bootstrap/5.3.0/css/bootstrap.min.css") }
            }
            body {
                div {
                    attrClass("container")
                    div {
                        attrClass("pb-2 mt-4 mb-3 border-bottom")
                        h1 { text("JFall 2013 Presentations - HtmlFlow") }
                    }
                    await { div, model: Observable<Presentation>, onCompletion ->
                        model
                            .doOnNext { 
                                presentationFragmentAsync
                                    .renderAsync(it)
                                    .thenApply { frag -> div.raw(frag) }
                            }
                            .doOnComplete { onCompletion.finish() }
                            .subscribe()
                    }
                }
            }
        }
    }.threadSafe()
\end{lstlisting}
\lstset{style=listingstyle}

HtmlFlow is an example of an \textit{internal} DSL that uses calls to HTML
builders as a way to emit HTML\@. It allows for the use of the
\texttt{Observable} class as a data model, which in turn allows for the definition of a
\texttt{Consumer<Presentation>} through the \texttt{doOnNext} method. Every
time a new value is emitted to the \texttt{Observable}, the \texttt{doOnNext}
method is called, emitting the HTML fragment for the \texttt{Presentation}
object to the client.

The Jstachio template engine, on the other hand, uses an \textit{external} DSL,
where the HTML template is defined in a separate file. The template engine uses
the \texttt{presentationItems} variable to define the data model, which is a
list of \texttt{Presentation} objects. This allows the template engine to
iterate over the list of \texttt{Presentation} objects and render the HTML for
each one.

As demonstrated, the HtmlFlow template enables PSSR by directly supporting the
use of the \texttt{Observable} class as a data model. However, not all
programmers are familiar with asynchronous programming models, which can make
implementing this solution challenging. In contrast, Jstachio defines templates
using standard HTML syntax with specific tags for the template engine,
resulting in a simpler and more accessible template definition. However,
because Jstachio templates rely on an \texttt{Iterable} interface,
they block the calling thread until the entire HTML content is rendered and
therefore do not support non-blocking PSSR\@.

% Mencionar que certas DSLs internas não suportam PSSR por 
% causa de HTML mal formado? (KotlinX)

In previous work \cite{PSSR-WISE2024}, Carvalho developed a benchmark
using Spring Webflux to evaluate the performance of HtmlFlow with suspendable
web templates. In this benchmark, HtmlFlow's performance was compared to that
of Thymeleaf using a reactive approach, as well as other template engines such
as the one shown in \autoref{lst:presentation-JStachio}. The results
demonstrated that HtmlFlow scaled efficiently up to 128 concurrent users,
achieving a throughput of up to 4,000 requests per second. In contrast, the
blocking template engines ceased to scale beyond 4 concurrent users, reaching a
maximum throughput of approximately 400 requests per second. These findings
indicate that engines which block the calling thread scale poorly and are not
suitable for PSSR\@.

% Expandir sobre virtual threads aqui?
With the advent of Java 21 and the introduction of \textit{Virtual Threads}, it
is now possible to execute code that would otherwise block the calling thread
in a scalable, non-blocking manner. This advancement enables the use of these
\textit{external} DSLs in non-blocking contexts. The objective of this work is
to evaluate the performance of these template engines when leveraging Virtual
Threads, and to determine whether they can be made viable for PSSR\@. We will
also compare the performance of this approach with existing solutions based on
reactive programming and Kotlin coroutines.

If the performance of these engines proves comparable to that of current
approaches, it would enable PSSR to be implemented using more familiar HTML
syntax, simplifying template development and adoption. Furthermore, template
engine developers seeking to support PSSR would no longer need to implement
complex asynchronous programming models to achieve non-blocking I/O; instead,
they could rely on the traditional approaches used for Server Side Rendering
(SSR).

>>>>

In this section, we examine the challenges of implementing Progressive
Server-Side Rendering (PSSR) in modern web applications, with a focus on the
limitations of current template engine designs. Our goal is to broaden the
range of options available for PSSR, particularly within JVM-based frameworks.

Reactive streams, such as those based on \texttt{Observable} (from RxJava) and
\texttt{Flow} (from the Java 9+ standard library), are abstractions that enable
progressive server-side rendering (PSSR) in environments utilizing
\textit{internal} DSLs. These abstractions provide a non-blocking, asynchronous
way to handle data streams, making them ideal for generating and sending HTML
content incrementally as data becomes available. In PSSR, the server does not
wait for the entire data model to be ready before beginning to render HTML.
Instead, it processes and streams each piece of data to the client as soon as
it arrives. Reactive types like \texttt{Observable<T>} or \texttt{Flow<T>}
facilitate this by representing data as a sequence of asynchronous events.
These streams emit values over time and support backpressure, allowing the
rendering engine to consume data at a manageable rate without overwhelming
system resources. A reactive stream might emit a sequence of
\texttt{Presentation} objects, each representing a talk in a conference
schedule. As each \texttt{Presentation} is emitted by the \texttt{Observable}
or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
corresponding HTML fragment and immediately flush it to the client. This
approach is demonstrated in \autoref{lst:presentation-observable}, where each
presentation is rendered asynchronously as it is emitted by an
\texttt{Observable}, and in \autoref{lst:presentation-flow}, which shows an
equivalent coroutine-based implementation using Kotlin's \texttt{Flow}. Both
examples highlight how internal DSLs can natively integrate with reactive types
to enable non-blocking, progressive rendering on the server side.

\lstset{style=listingstyle}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={\textit{HtmlFlow} presentation template with \texttt{Observable}},
    label={lst:presentation-observable}
]
await { div, model: Observable<Presentation>, onCompletion ->
    model
        .doOnNext { 
            presentationFragmentAsync
                .renderAsync(it)
                .thenApply { frag -> div.raw(frag) }
        }
        .doOnComplete { onCompletion.finish() }
        .subscribe()
}
\end{lstlisting}
\hfil
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    label={lst:presentation-flow},
    caption={\textit{HtmlFlow} presentation template with \texttt{Flow}},
]
suspending { model: Flow<Presentation> ->
    model.collect {
        presentationFragmentAsync
            .renderAsync(it)
            .thenApply { frag -> raw(frag) }
    }
}
\end{lstlisting}

By contrast, template engines that use \textit{external} DSLs—such as JStachio,
Thymeleaf, or Handlebars—typically define templates within static HTML
documents using custom markers, and rely on blocking interfaces like
\texttt{Iterable} or \texttt{List}. These interfaces require the entire data
model to be materialized before rendering can begin, which blocks server
threads during template expansion and significantly limits scalability under
high concurrency. These interfaces require the entire data model to be
materialized before rendering can begin, which blocks server threads during
template expansion and significantly limits scalability under high concurrency.
Some reactive libraries, such as RxJava, provide bridging mechanisms like
\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
to be exposed as Iterable by blocking the thread until new items are available.
While useful for compatibility with traditional APIs, this approach
reintroduces blocking behavior and undermines the benefits of non-blocking
I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
illustrates this model using a JStachio template, where the engine performs a
blocking loop over \texttt{presentationItems}.

\vspace{1cm}

\lstset{style=listingstyle}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{JStachio}},
  label={lst:presentation-jstachio}
]
{{#presentationItems}}
<div class="card mb-3 shadow-sm rounded">
    <div class="card-header">
        <h5 class="card-title">
            {{title}} - {{speakerName}}
        </h5>
    </div>
    <div class="card-body">
        {{summary}}
    </div>
</div>
{{/presentationItems}}
\end{lstlisting}

Despite these performance limitations, external DSLs remain popular due to
several advantages:

\begin{enumerate}
    \item \emph{Separation of Concerns}: HTML templates are decoupled from application logic, enabling front-end developers to contribute without modifying back-end code.
    \item \emph{Cross-Language Compatibility}: External DSLs are portable across languages and frameworks, easing integration in multi-language environments.
    \item \emph{Familiarity}: Many developers are comfortable with HTML syntax, lowering the barrier to entry and improving maintainability.
\end{enumerate}

These strengths make external DSLs appealing—even when they come at the cost of
blocking, synchronous rendering. However, this tradeoff becomes critical under
high concurrency, where blocking threads severely degrades throughput. Emerging
features in the Java ecosystem, particularly \textit{virtual threads}
introduced in Java 21 as part of Project Loom, offer a promising solution to
this challenge. Virtual threads drastically reduce the overhead of blocking
operations by decoupling thread execution from OS-level threads. As a result,
engines that rely on blocking interfaces—like those used in
external DSLs—can potentially achieve scalability levels that approach those of
non-blocking, asynchronous engines.

% Incluir este parágrafo?
% In previous work \cite{PSSR-WISE2024}, Carvalho benchmarked the performance of
% HtmlFlow (an internal DSL) using suspendable templates within a Spring WebFlux
% application. HtmlFlow scaled efficiently to 128 concurrent users and delivered
% up to 4,000 requests per second. In contrast, blocking engines such as JStachio
% saturated at just 4 concurrent users, maxing out at around 400 requests per
% second—highlighting the scalability limitations of blocking models.This work
% builds upon those findings by evaluating whether virtual threads can close this
% performance gap for blocking template engines. Specifically, we aim to
% determine whether external DSLs—traditionally unsuitable for PSSR due to their
% reliance on blocking APIs—can become viable for progressive rendering when
% powered by virtual threads.

If successful, this approach would enable developers to reap the ergonomic
benefits of external DSLs while maintaining scalable performance, without
adopting complex asynchronous paradigms. It would also simplify the
implementation of PSSR across a broader range of frameworks and engines, making
the technique more accessible to mainstream web development.
